---
title: "Categorical models and model comparison"
editor: source
freeze: true
---

## Adding categorical predictors

```{r}
library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(ggdist)
library(brms)
library(tidybayes)
library(tidyr)
theme_set(theme_minimal())

penguins <- penguins |>
  filter(!is.na(sex),
         species == "Chinstrap") 

ggplot(penguins, aes(bill_depth_mm, 
                     bill_length_mm, 
                     color = sex)) +
  geom_point()
```

### The model

-   Say we want to model `bill_length` as a function of both `bill_depth` and `sex`.
-   There are different approaches to adding a categorial predictor to a model, sometimes called "indicator" (or "dummy") variables, or "index" variables. McElreath comes down strongly in favor of using "index" variables. For more discussion, see [here](https://bookdown.org/content/4857/the-many-variables-the-spurious-waffles.html#categorical-variables) and links therein.

### The code

### Fitting the model

First we can fit a model that will just fit different *intercepts* for the different species:

```{r, eval = F}


depth_length_sex_brm <- brm(
  family = gaussian,
  data = penguins,
  formula = bill_depth_mm ~ 0 + bill_length_mm + sex,
  prior = c(
      prior(normal(0, 5), class = b),
      prior(uniform(0, 20), class = sigma, ub = 20)),
  iter = 1000
)

```

```{r, eval = F, echo = F}

save(depth_length_sex_brm, file = "bill_depth_sex_lm.RDS")

```

```{r, echo = F}
load("bill_depth_sex_lm.RDS")
```

### Interpreting the model

Model diagnostics:

```{r}

summary(depth_length_sex_brm)

plot(depth_length_sex_brm)

```

Comparing the intercepts for `sex`:

```{r}

as_draws_df(depth_length_sex_brm) %>% 
  mutate(diff_fm = b_sexfemale - b_sexmale) %>% 
  pivot_longer(cols = c(b_sexfemale:sigma, diff_fm)) %>% 
  group_by(name) %>% 
  mean_qi(value, .width = .89)

```

### Model comparison

See chapter 7 and the bookdown book [here](https://bookdown.org/content/4857/ulysses-compass.html) for more on model comparison criteria.

```{r}


depth_length_sex_brm <- depth_length_sex_brm |> add_criterion(criterion = c("waic", "loo"))

depth_length_sex_brm$criteria

```

Compare this to our old model, with no term for sex:

```{r, echo = F}
load("bill_depth_lm.RDS")
```

```{r}


depth_length_brm <- depth_length_brm |> add_criterion(criterion = c("waic", "loo"))

depth_length_brm$criteria


```
