---
title: "Components of a model"
editor: source
freeze: true
---

Based on the book section [4.2. A language for describing models](https://learning.oreilly.com/library/view/statistical-rethinking-2nd/9780429639142/xhtml/06_Contents.xhtml#sec4_2).

## Variables

### Data

Observable things.

### Parameters

Not observable or known.

## "Joint generative model"

We define each variable either in terms of a probability distribution, or in terms of its relationship to the other variables.

Taken together, this system of definitions is what McElreath calls a "joint generative model".

## Penguins example

Let's model `bill_length` for Chinstrap penguins.
First, we won't use any predictor variables - just looking for the mean.

```{r}

library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(ggdist)
theme_set(theme_minimal())

penguins <- penguins |>
  filter(!is.na(sex)) |>
  filter(species == "Chinstrap") 

ggplot(penguins, aes(bill_depth_mm)) +
  geom_dotsinterval()

```

### Defining the model

$\text{billdepth}_i \sim N(\mu, \sigma)$

This just says that bill depth is normally distributed with some mean and some standard deviation.
We can observe bill depth (data), but we don't know the mean or standard deviation (parameters).

For the parameters, we need *priors*.

$\mu \sim N(50, 15)$

This is to say that $\mu$ is some number drawn from a normal distribution centered on 50:

```{r}

data.frame(mu = rnorm(1000, mean = 20, sd = 6)) |>
  ggplot(aes(x = mu)) +
  geom_dotsinterval() +
  ggtitle("Simulations from prior for mu")

```

\$\sigma \sim \text{Uniform}(0, 20)\$

And here we're saying that we figure $sigma$ is potentially uniformly distributed ranging from 0-20.
Standard deviations have to be positive, and otherwise this is a very broad range.

```{r}

data.frame(sigma = runif(1000, 0, 20)) |>
  ggplot(aes(x = sigma)) +
  geom_dotsinterval() +
  ggtitle("Simulations from prior for sigma")

```

### Exploring the generative model

```{r}

sample_mu <- rnorm(1000, 20, 6)
sample_sigmas <- runif(1000, 0, 20)

data.frame(simulated_bill_depths = 
             rnorm(1000, 
                   sample_mu,
                   sample_sigmas)) |>
  ggplot(aes(simulated_bill_depths)) +
  geom_dotsinterval() +
  ggtitle("Simlated bill depths from priors")
```

This is our simulation of expected bill depths *before explicitly taking into account the data* (although we did look at the density plot before specifying the priors).

### Taking into account the data

See [here](https://bookdown.org/content/4857/geocentric-models.html#finding-the-posterior-distribution-with-quap-brm.) for translations of `rethinking` code to `brms`.

```{r, eval = F}
library(brms)

bill_depth_brm <-
  brm(
    family = gaussian,
    bill_depth_mm ~ 1,
    data = penguins,
    prior = c(
      prior(normal(20, 6), class = Intercept),
      prior(uniform(0, 20), class = sigma, ub = 20)
    ),
    iter = 1000
  )

```

```{r, eval = F, echo = F}

save(bill_depth_brm, file = "bill_depth_only.RDS")
```

```{r, eval = T, echo = F}
library(brms)

load("bill_depth_only.RDS")

```

```{r}
plot(bill_depth_brm)

```

```{r}

summary(bill_depth_brm)

```

### Exploring the posterior

```{r}

posterior <- as_draws_df(bill_depth_brm)

head(posterior)

```

```{r}

ggplot(posterior, aes(Intercept)) +
  stat_dotsinterval()


ggplot(posterior, aes(sigma)) +
  stat_dotsinterval()
```

```{r}

posterior_summary(bill_depth_brm)

```

## Penguins with a predictor

OK, now let's model bill_depth as a function of bill_length.

```{r}

ggplot(penguins, aes(bill_length_mm, 
                     bill_depth_mm)) +
  geom_point()

```

### Specifying the model

$\text{billdepth}_i \sim N(\mu, \sigma)$

Here, again, we say that bill depth is normally distributed with some mean and standard deviation.

$\mu_i = \alpha + \beta(\text{billlength}_i = \text{mean}(\text{billlength}))$

This addition makes it into a linear model.
Instead of estimating $\mu$ from the data, we say that the mean of bill depth varies with bill length as linear function with an intercept $\alpha$ and a slope $\beta$.

$\alpha$ and $\beta$ are now additional parameters that we will estimate and therefore need to set priors for.

$\alpha \sim N(20, 6)$

$\beta \sim N(0, 5)$

$\sigma \sim \text{Uniform}(0, 20)$

### Exploring the generative model

### Fitting the model (with data)

### Exploring the posterior
