---
title: "Fitting and interpreting interaction models"
editor: source
freeze: true
---

## New dataset: this time with mice!

```{r, eval = F, echo = T}

install.packages('ratdat')

```

```{r}
library(ratdat)
library(dplyr)
library(ggplot2)
library(ggdist)
library(brms)
library(tidybayes)
library(tidyr)
theme_set(theme_minimal())

complete_timestamp <- ratdat::complete |>
  mutate(date = as.Date(paste(month, day, year, sep = "-"), format = "%m-%d-%Y"),
         scientific_name = paste(genus, species, sep = "_")) |>
  filter(!is.na(weight),
         !is.na(hindfoot_length),
         species_id %in% c("DM", "PP")) |>
  mutate(weight = scale(weight),
         numeric_date = as.numeric(date))


ggplot(complete_timestamp, aes(date, weight, color = species_id)) +
  geom_point(alpha = .1) +
  geom_smooth(method = "lm")


```

## The models

-   Say we want to model `weight` as a function of `time`, and we are curious how species affects this relationship

### Weight and species (but no interaction)

```{r, eval = F, echo = T}

weight_species_brm <- brm(
  family = gaussian,
  data = complete_timestamp,
  formula = weight ~ 0 + numeric_date + species_id,
  prior = c(
    prior(normal(0, 5), class = b),
    prior(uniform(0, 20), class = sigma, ub = 20)),
  iter = 4000,
  file = "weight_species_brm.RDS"
)
```

### Weight and species with interaction

```{r, eval = F, echo = T}
weight_species_interaction_brm <- brm(
  data = complete_timestamp, 
  family = gaussian,
  bf(weight ~ 0 + a * numeric_date, 
     a ~ 0 + species_id, 
     nl = TRUE),
  prior = c(prior(normal(1, 0.1), class = b, coef = DM, nlpar = a),
            prior(normal(1, 0.1), class = b, coef = PP, nlpar = a),
            prior(exponential(1), class = sigma)),
  iter = 4000)
```

```{r, echo = F, eval = T}
weight_species_brm <- readRDS("weight_species_brm.RDS")
weight_species_interaction_brm <- readRDS("weight_species_interaction_brm.RDS")

```

```{r}
summary(weight_species_interaction_brm)
```

### Interpreting the model

Model diagnostics:

```{r}

summary(depth_length_sex_brm)

plot(depth_length_sex_brm)

```

Comparing the intercepts for `sex`:

```{r}

as_draws_df(depth_length_sex_brm) %>% 
  mutate(diff_fm = b_sexfemale - b_sexmale) %>% 
  pivot_longer(cols = c(b_sexfemale:sigma, diff_fm)) %>% 
  group_by(name) %>% 
  mean_qi(value, .width = .89)

```

### Model comparison

See chapter 7 and the bookdown book [here](https://bookdown.org/content/4857/ulysses-compass.html) for more on model comparison criteria.

```{r}


depth_length_sex_brm <- depth_length_sex_brm |> add_criterion(criterion = c("waic", "loo"))

depth_length_sex_brm$criteria

```

Compare this to our old model, with no term for sex:

```{r, echo = F}
load("bill_depth_lm.RDS")
```

```{r}


depth_length_brm <- depth_length_brm |> add_criterion(criterion = c("waic", "loo"))

depth_length_brm$criteria


```
