---
title: "Fitting and interpreting interaction models"
editor: source
freeze: true
---

## New dataset: this time with mice!

### The Portal data

[The Portal Project](https://portal.weecology.org/): 40 years of population surveys of rodents in southeastern Arizona.

We'll focus on abundance trends in two species, *Dipodomys spectabilis* and *Chaetodipus penicillatus*.
[More on these species.](https://portalcasting.netlify.app/profiles)

### Getting the Portal teaching data

```{r, eval = F, echo = T}

install.packages('ratdat')

```

### Plotting trends in DS and PP

```{r}
library(ratdat)
library(dplyr)
library(ggplot2)
library(ggdist)
library(brms)
library(tidybayes)
library(tidyr)
theme_set(theme_minimal())

complete_annual <- ratdat::complete |>
  group_by(year, species_id) |>
  filter(species_id %in% c("DS", "PP")) |>
  tally() |>
  ungroup() |>
  mutate(n = scale(n))

ggplot(complete_annual, aes(year, n, color = species_id)) +
  geom_point()



```

## Fitting models

-   Say we want to model `abund` as a function of `time`, and we are curious how species affects this relationship

### abund and species (but no interaction)

```{r, eval = F, echo = T}

abund_species_brm <- brm(
  family = gaussian,
  data = complete_annual,
  formula = n ~ 0 + year + species_id,
  iter = 4000, 
  warmup = 1000, 
  chains = 4, 
  cores = 4,
  seed = 1977,
  file = "abund_species_brm",
  control = list(max_treedepth = 15)
)

```

### abund and species with interaction

To fit an interaction model with index (rather than indicator) variables, use the brms nonlinear syntax.
[More here!](https://bookdown.org/content/4857/conditional-manatees.html#building-an-interaction)

```{r, echo = T}
abund_species_interaction_brm <-  brm(data = complete_annual, 
      family = gaussian,
      bf(n ~ 0 + a + b * year, 
         a ~ 0 + species_id,
         b ~ 0 + species_id,
         nl = TRUE),
      iter = 4000, 
      warmup = 1000, 
      chains = 4, 
      cores = 4,
      control = list(max_treedepth = 15),
      seed = 8,
      file = "abund_species_interaction_brm")
```

```{r, echo = F, eval = T}
abund_species_brm <- readRDS("abund_species_brm.RDS")
abund_species_interaction_brm <- readRDS("abund_species_interaction_brm.RDS")

```

## A note on priors

```{r}
default_prior(abund_species_brm)
default_prior(abund_species_interaction_brm)
```

## Model results

### No interaction

```{r}
summary(abund_species_brm)
plot(abund_species_brm)
```

```{r}

abund_species_brm_predictions <- complete_annual |>
  add_epred_draws(abund_species_brm, ndraws = 50)

ggplot(abund_species_brm_predictions, aes(year, n, color = species_id, fill = species_id)) +
  geom_point() +  
  stat_lineribbon(aes(y = .epred), alpha = .1) 

```

### Interaction

```{r}

summary(abund_species_interaction_brm)
plot(abund_species_interaction_brm)

```

```{r}

abund_species_interaction_brm_draws <- tidy_draws(abund_species_interaction_brm)

ggplot(abund_species_interaction_brm_draws, aes(b_b_species_idDS - b_b_species_idPP)) +
  geom_histogram()

abund_species_interaction_brm_predictions <- complete_annual |>
  add_epred_draws(abund_species_interaction_brm, ndraws = 50)

ggplot(abund_species_interaction_brm_predictions, aes(year, n, color = species_id, fill = species_id)) +
  geom_point() +  
  stat_lineribbon(aes(y = .epred), alpha = .1) 

```
